From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-02 Pi+/paper/finite.tex
Redo Lehmer algorithm example

\todo{typify}
\begin{definition}
  Given $\sigma : \Aut[\Fin[n]]$, the inversion count of $k: \Fin[n]$ is given by
  $|\Set{j < i | \sigma(j) > \sigma(i)}|$.
\end{definition}

It turns out that knowing just the inversion counts for all the elements, one can reconstruct the starting permutation.
Also, observe that the inversion count for element $i$ is guaranteed to be smaller than $i$, thus fitting in the $i$-th
place of a Lehmer code tuple. As an example, consider the following tabulated presentation of the permutation:

\todo{fix this figure}

\[
  \sigma =
  \begin{array}{ccccccccccccccc}
    | & 0      & | & 1      & | & 2      & | & 3      & | & 4      & | \\
    \hline                                                             \\
    | & \el{2} & | & \el{1} & | & \el{4} & | & \el{0} & | & \el{3} & | \\
    \hline                                                             \\
  \end{array}
\]

%  0 1 2 3 4
% -----------
% |2|0|1|4|3|
% -----------

The inversion count for $\el{0}$ is 0 (because there are no smaller elements at all), for $\el{1}$ is 0 (because of
$\el{0}$ appearing after), for $\el{2}$ is 2 (because of $\el{0}$ and $\el{1}$), for $\el{3}$ is 0 (because it comes
last in the sequence), and for $\el{4}$ is 1 (because of $\el{3}$). Thus, the Lehmer code for the permutation $\sigma$
is the 5-tuple $l = (0, 1, 2, 0, 1)$.

To construct the presentation of the permutation from the Lehmer code, written using adjacent transpositions, we perform
an algorithm similar to \emph{insertion sort}. Starting from the left-most position of the tuple $l$, we'll read the
value $v$, insert the new element at the end of the newly created list, and shift it backward $v$ places.

\begin{center}
  \begin{tabular}{c|p{0.75\linewidth}}
    (0, 0, 2, 0, 1)               & We start from an empty list $[]$                                                                 \\
    (\highlight{{0}}, 0, 2, 0, 1) & We read 0 as the left-most value from $l$. Thus, we append the element $\el{0}$ to our
    list, getting $[\el{0}]$. The element is shifted $0$ places, so it remains in the
    same place.                                                                                                                      \\
    (0, \highlight{{0}}, 2, 0, 1) & Then, similarly, we read another 0 for the element $\el{1}$, append it to the
    list getting $[\el{0}, \el{1}]$, and don't shift it either.                                                                      \\
    (0, 0, \highlight{{2}}, 0, 1) & We read 2 for the next the element $\el{2}$ - we append $\el{2}$ to our list, getting
    $[\el{0}, \el{1}, \el{2}]$, and shift it 2 places right, which results in a list $[\el{0}, \el{2}, \el{1}]$
    \todo{Typeset it nicely, with arrows showing the shifting}.                                                                      \\
    (0, 0, 2, \highlight{{0}}, 1) & Then we read 0 - appending $\el{3}$ and not shifting, getting $[\el{0}, \el{2}, \el{1}, \el{3}]$ \\
    (0, 0, 2, 0, \highlight{{1}}) & Finally, reading 1 for element $\el{4}$ - appending $\el{4}$ to the list and shifting it
    one place right results in the final list $[\el{0}, \el{2}, \el{1}, \el{4}, \el{3}]$                                             \\
  \end{tabular}
\end{center}
\todo{figure}

Writing formally, to turn a code into a permutation written in the Coxeter presentation, we use $\immersion$:

@@

As described above, the number $r$ at the position $k$ in the tuple describes how many inversions the element $\el{k}$
has. Thus, we need to perform $r$ many adjacent transpositions to get to the desired position, which is given by
$(\suc[n] - r) \downf r$.

>>>>

\todo{typify}
\begin{definition}
  Given a permutation $\sigma : \Aut[\Fin[n]]$, the inversion count of $k: \Fin[n]$ is given by
  \[ |\Set{j < i | \sigma(j) > \sigma(i)}|. \]
\end{definition}

It turns out that from knowing just the inversion counts for all the elements, one can reconstruct the starting
permutation. Also, observe that the inversion count for element $i$ is guaranteed to be smaller than $i$, thus fitting
in the $i$-th place of a Lehmer code tuple. As an example, consider the following tabulated presentation of a
permutation of $\Fin[5]$.

\[
  \sigma =
  \begin{pmatrix}
    0      & 1      & 2      & 3      & 4      \\
    \el{2} & \el{1} & \el{4} & \el{0} & \el{3} \\
  \end{pmatrix}
\]

The inversion count for $\el{0}$ is 0 (because there are no smaller elements at all), for $\el{1}$ is 1 (because of
$\el{0}$ appearing after), for $\el{2}$ is 2 (because of $\el{0}$ and $\el{1}$), for $\el{3}$ is 0 (because it comes
last in the sequence), and for $\el{4}$ is 2 (because of $\el{3}$ and $\el{0}$). Thus, the Lehmer code for the permutation $\sigma$
is the 5-tuple $l = (0, 1, 2, 0, 2)$.

To decode the permutation back from this Lehmer code, we perform an algorithm similar to \emph{insertion sort}. The
element of the Lehmer code being currently processed is highlighted in the left column of the table below. Starting from
a sorted list, the element at index $k$ has to be given $l[k]$ inversions. Because of the the invariant that all the
elements before newly processed one are smaller than it, the proper number of inversions is created by simply shifting
the element $l[k]$ places left.

\begin{center}
  \begin{tabular}{l|r}
    (\highlight{{0}}, 1, 2, 0, 2) & $[\highlightAlt{\el{0}}, \el{1}, \el{2}, \el{3}, \el{4}]$ \\
    (0, \highlight{{1}}, 2, 0, 2) & $[\highlightAlt{\el{1}}, \el{0}, \el{2}, \el{3}, \el{4}]$ \\
    (0, 1, \highlight{{2}}, 0, 2) & $[\highlightAlt{\el{2}}, \el{1}, \el{0}, \el{3}, \el{4}]$ \\
    (0, 1, 2, \highlight{{0}}, 2) & $[\el{2}, \el{1}, \el{0}, \highlightAlt{\el{3}}, \el{4}]$ \\
    (0, 1, 2, 0, \highlight{{2}}) & $[\el{2}, \el{1}, \highlightAlt{\el{4}}, \el{0}, \el{3}]$ \\
  \end{tabular}
\end{center}

Writing formally, to turn a code into a permutation written in the Coxeter presentation, we define a function
$\immersion$. As described above, the number $r$ at position $k$ in the tuple describes how many inversions the element
$\el{k}$ has. Thus, we need to perform $r$ many adjacent transpositions to get to the desired position, which is given
by $(\suc[n] - r) \downf r$.

@@
